Kindly Bent to Free Us

### 2019-11-12 kindly/intro.tex
Adapt to the PLDI format.

The most closely related contenders in this design space are Linear Haskell
\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth called LH,
Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
Compared to LH, the goals and means are
similar as both systems permit abstraction over the number of uses of
values and retain type inference, but the details are quite different.
\begin{enumerate}
\item LH's multiplicities are either linear or unrestricted whereas
  we also permit affine values.
\item Our system supports mutable and immutable borrows, hence the
  need for affinity.
\item Our multiplicities are directly attached to the type of the
  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
  denotes the type of a \emph{single-use function} that can be called
  just once, whereas the multiplicities in LH choose
  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
  latter is a function that promises to \emph{use its argument exactly
    once}.
\item Similar to Quill, we make use of multiplicity contraints. We
  rely on kind subsumption to simplify the inferred types.
\end{enumerate}
See \cref{sec:related-work} for further in-depth discussion of these
and other related works.
\subsection{First examples}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.45\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  File.write &!h "Hello ";
  File.write &!h "world!";
  File.close h
\end{lstlisting}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

As a first, well-known example we consider a simplified API for
writing files shown in \cref{fig:writing-files-api}.  It introduces a
linear abstract type \lstinline/File.t/. A call like
\lstinline/fopen "foo"/ returns a linear handle to a newly created
file, which \emph{must} be released later on through \lstinline/close/
as shown in \cref{fig:writing-files-example}. Failing to do so is a
static type error.  To write to the file, we must take a (mutable)
borrow \lstinline/&!h/ of the handle and pass it to the
\lstinline/write/ function. Mutable borrows are classified as affine:
they must not be duplicated, but there is no requirement to use
them. This affinity shows up in the annotation \lstinline/-{aff}>/ of the second arrow in
the type of \lstinline/write/: a partial application like
\lstinline/write &!h/ captures the affine borrow and hence the
resulting function must also be handled in an affine way. That is, it
would be an error to write
\begin{lstlisting}
  let w = File.write &!h in w "Hello "; w "world!" (* type error *)
\end{lstlisting}
as we would be using the affine closure twice. The remaining arrows in
the API are unrestricted and we write \lstinline/->/ instead of the
explicitly annotated \lstinline/-{un}>/.

@@

\begin{itemize}
\item A type system that encodes linearity and affinity with
  borrowing. The type system is polymorphic with respect to types as
  well as kinds that express constraints on the number of uses of a value.
\item An extension of the HM(X) framework
  \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
  multiplicity constraints that supports full, principal type inference.
\item Full account of the metatheory for type inference and type soundness.
\item Constraint simplification techniques.
\item Automatic inference of regions for borrows.
\item An implementation of the type inference algorithm including all
  constraint simplification and extended with pattern matching. A
  webfront for the algorithm is available, its URL is given in the
  non-anonymous supplementary material, and we plan to submit it for
  artifact evaluation.
\end{itemize}

>>>>

The most closely related contenders in this design space are Linear Haskell
\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth called LH,
Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
Compared to LH, the goals and means are
similar as both systems permit abstraction over the number of uses of
values and retain type inference, but the details are quite different.
\begin{enumerate}
\item LH's multiplicities are either linear or unrestricted whereas
  we also permit affine values.
\item Our system supports mutable and immutable borrows, hence the
  need for affinity.
\item Our multiplicities are directly attached to the type of the
  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
  denotes the type of a \emph{single-use function} that can be called
  just once, whereas the multiplicities in LH choose
  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
  latter is a function that promises to \emph{use its argument exactly
    once}.
\item Similar to Quill, we make use of multiplicity contraints. We
  rely on kind subsumption to simplify the inferred types.
\end{enumerate}
See \cref{sec:related-work} for further in-depth discussion of these
and other related works.
\subsection{First examples}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \vspace{-15pt}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  File.write &!h "Hello ";
  File.write &!h "world!";
  File.close h
\end{lstlisting}
    \vspace{-10pt}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \vspace{-10pt}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

As a first, well-known example we consider a simplified API for
writing files shown in \cref{fig:writing-files-api}.  It introduces a
linear abstract type \lstinline/File.t/. A call like
\lstinline/fopen "foo"/ returns a linear handle to a newly created
file, which \emph{must} be released later on through \lstinline/close/
as shown in \cref{fig:writing-files-example}. Failing to do so is a
static type error.  To write to the file, we must take a (mutable)
borrow \lstinline/&!h/ of the handle and pass it to the
\lstinline/write/ function. Mutable borrows are classified as affine:
they must not be duplicated, but there is no requirement to use
them. This affinity shows up in the annotation \lstinline/-{aff}>/ of the second arrow in
the type of \lstinline/write/: a partial application like
\lstinline/write &!h/ captures the affine borrow and hence the
resulting function must also be handled in an affine way. That is, it
would be an error to write
\begin{lstlisting}
let w = File.write &!h in w "Hello "; w "world!" (*error*)
\end{lstlisting}
as we would be using the affine closure twice. The remaining arrows in
the API are unrestricted and we write \lstinline/->/ instead of the
explicitly annotated \lstinline/-{un}>/.

@@

\begin{itemize}
\item A type system that encodes linearity and affinity with
  borrowing. The type system is polymorphic with respect to types as
  well as kinds that express constraints on the number of uses of a value.
\item An extension of the HM(X) framework
  \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
  multiplicity constraints that supports full, principal type inference.
\item Full account of the metatheory for type inference and type soundness.
\item Constraint simplification techniques.
\item Automatic inference of regions for borrows.
\item An implementation of the type inference algorithm including all
  constraint simplification and extended with pattern matching. A
  web implementation of the typechecker is available, its URL is given in the
  non-anonymous supplementary material, and we plan to submit it for
  artifact evaluation.
\end{itemize}
